Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-04 manuscript/manuscript.tex
minor ms changes
ms updates

Previous work in population genomics has established that negative selection
against against new deleterious variants introduced by mutation profoundly
shapes large-scale patterns of variation across the genome. However, the
mathematical models that describe this process only consider variation so
deleterious that it cannot drift up in frequency and ultimately fix in the
population. In reality, many new mutations are only mildly deleterious, and can
fix in the population if random perturbations from genetic drift or selection
at neighboring sites (genetic draft) overpowers the selection against them.
Despite the large role negative selection plays across the human genome, no
existing evolutionary models consider the joint effects of mutation, selection,
and genetic drift and drift across the genome. Consequently, population genetic
models cannot describe the ultimate fate of every new variant ---eventual
extinction or fixation in the population--- and thus cannot connect to observed
rates of molecular evolution.

whole-genome variance models.
These are flux models.

This is because, in the very long run, all mutations must be accounted for.

Here, we introduce a new whole-genome general mutation-selection-draft-drift
(MSD$^2$) model that extends and implements previous theory by
\textcite{Santiago2016-mu}.

Unlike classic mutation-selection models which cannot accurately capture the
dynamics of weakly deleterious mutation, MSD$^2$ models can, and consequently
can provide estimates of the distribution of selection coefficients of new
mutations. Additionally, because our model incorporates drift and draft, our
model makes explicit predictions about the substitution rate of weakly
deleterious mutations and allows us to test XXX.

@@

If the human evolutionary process were strongly impacted by Hill--Robertson
Interference, we should expect that substitution rates are highest in regions
experiencing the most selection, which are regions when $B(x)$ is the lowest.
However, our model

Evolutionary processes describe how genetic variation is introduced and
ultimately converted to substitutions or lost from the population. Our model 
makes predictions about the eventual outcome of weakly deleterious mutation 
that seem to roughly match observations.

Note on the importance of selection in humans.
NOTE about the importance of calibration

emphasis on whole genome models

Eukaryotes genomes are under continual selectionary pressure to remove the
continual influx of deleterious mutations at conserved regions. This purifying
selection process keeps mutations that reduce fitness at low frequency in the
population, and reduces genetic diversity at sites linked to genomic regions
that are predominantly under purifying selection such as coding sequences and
regions that have been shown to be phylogenetically conserved between species.
Evolutionary geneticists have developed a series of mathematical models of this
process, to predict quantities such as the equilibrium level of segregating
deleterious mutations (\emph{genetic load}), and the reduction in neutral
genetic diversity at linked sites due to a faster rate of coalescence (the $B$
reduction factor). Statistical models of the purifying selection process fit to
human genomic data closely match observed patterns of genetic variability along
the genome, explaining around 60\% of the genetic variability at the megabase
scale \parencite{McVicker2009-ax,Murphy2022-sj}.

These evolutionary genetics models only consider the impact of new mutations
that are sufficiently deleterious that they are guaranteed to be eliminated
eventually by selection. When mutations are strongly deleterious, the impact of
random genetic drift and random perturbations caused by selection at linked
neighboring sites (genetic draft) on these mutations on their fates can be
ignored. Under these conditions, mathematical models can assume an stationary
balance between the input of new mutations entering the population at rate
$\mu$ per basepair, per generation, and their rate of loss determined by the
relative reduction in heterozygous fitness, $1-s$. Under these
mutation-selection balance models, the equilibrium genetic load and the
reduction in coalescent time are approximately determined by the ratio
$\nicefrac{\mu}{s}$. However, estimates of the distribution of fitness effects
suggest many new mutations are weakly deleterious, such that their dynamics can
no longer ignore the influence of genetic drift and draft.

At every spot in the genome the processes of mutation-selection-draft-drift
against alleles deleterious in most environments reduce pairwise diversity. We
can imagine some unobserved reduction map, $B(x)$ across the genome. Here we
derive theory for this map under these processes and show how statistical
models can be fit using this theory.

To understand the this process and make predictions about the equilibrium level
of deleterious mutations segregating in the population, evolutionary
geneticists have used models of mutation-selection balance. These models

Non-equilibrium models of deleterious mutation flux.

Generalized background selection

These models

\section*{Results}

%\begin{figure}[!htb]
%  \centering
%  \includegraphics{figures/experimental_1.pdf}
%  \caption{}
%  \label{fig:figure-1}
%\end{figure}

@@

It is worth considering why this model performs well for models of negative
selection. The fitness variance equation, $V(x) = (U-2R)s$ tells us that under
MSD$^$2 models, the long-run variance is set by the balance of mutation and
fixation. Whether a particular basepair's behavior follows this equation
depends upon whether the long-run average is close to ``typical" for that
basepair at a given time. Good things fix so quickly, that we're left with a
constant churning of new variation

>>>>

@@

The continual influx of new mutations into populations is the ultimate source
of all adaptations, but the vast majority of mutations either do not have an
effect on fitness or are deleterious. Negative selection works to eliminate
such deleterious mutations from the population, which is essential to
maintaining the integrity of information in the genome to maintain a lineage.
Mutations under negative selection are kept at low frequencies (Haldane?),
contributing to human traits (XXX) and diseases. As with most selection in the
genome, this process perturbs the allele frequencies of neighboring linked
variation. This reduces genetic variability around conserved segments and
creates large-scale spatial patterns in genetic diversity. These patterns are
mediated by the deleterious mutation rate, the strength of selection in
conserved segments, and the spatial distribution of recombination rates and
conserved segments. Since genome-wide recombination maps and putatively
conserved segments are available for many species, researchers have combined
these with population genetic theory to estimate the mutation rate and
selection coefficients of deleterious mutations, as well as the overall
reduction in genetic variability due to the negative selection process.

Classic Background Selection (BGS) theory is the predominant model for the
reduction in linked neutral diversity used by statistical approaches to
quantify the impacts of negative selection. However, this model makes some
assumptions for mathematical tractability that could distort inferences about
the negative selection process. First, since fixation probabilities ultimately
depend on the product of the deleterious selection coefficient (s) and
population size (N), the efficacy of negative selection should depend on
demography. Unfortunately, accommodating demography into theoretic negative
selection models remains an open, difficult problem (XXX). Second, building off
classic models of mutation-selection balance (Crow 1970, Kimura and Maruyama
196), the Background Selection (XXX) model assumes that new mutations are
sufficiently deleterious that they are invariably driven to loss. Under this
assumption, the effect of selection is well-approximated by simply rescaling
the neutral coalescent by a reduction factor known as B (XXX). However, this is
only appropriate in a limited domain of parameters relevant to the negative
selection process (XXX). Because the BGS model cannot accommodate the
possibility of fixation for mutations with $Ns \ll 1$, diversity is incorrectly
predicted to fall to zero as the strength of selection diminishes. This weak
selection problem is related to the wider problem of modeling the substitution
rate of deleterious mutation (i.e. the “ratchet” rate). Third, the BGS model
assumes segments under purifying selection are sufficiently far apart that
there is no selective interference between them. This is a matter of degree,
since under sufficiently high mutation rates or with little recombination,
selective interference can occur with a segment.

In this work, we extend another class of linked selection models that derive
from quantitative genetics to address limitations of the classic BGS model.
These models consider how polygenic fitness variation increases the rate of
stochastic allele frequency change as alleles become randomly linked to fitness
backgrounds. While these models can theoretically accommodate polygenic fitness
variation from any source as long as its rate of change is not too rapid, we
focus specifically on a deleterious-mutations-only model from Santiago and
Caballero (2016). This model is identical to the classic BGS model when
selection against deleterious mutations is strong, but it also correctly
predicts the reduction in diversity and substitutions rates of weakly
deleterious mutations by accurately modeling levels of genic fitness variation.
We extend Santiago and Caballero’s (hereafter the SC16) model so that it can be
used to model genome-wide patterns of diversity under negative selection
processes, and develop Python software to use this theoretic approach to fit
genome-wide models of negative selection. Using forward simulations, we show
this model leads to more accurate DFE estimates under weak selection. While
this new model accommodates the possibility that a given reduction B could be
due to weaker selection, in applying our composite-likelihood method to human
population genomic data, we confirm earlier findings that strong negative
selection at a small fraction of sites fits human data across populations best.
Finally, we show that while this model solves some limitations of the classic
BGS model, simulations reveal another mysterious defect: a poor fit around
$Ns=1$. We show that this can be partially corrected by incorporating the effects
of selective interference on genic variation, but the remaining discrepancy is
due to the build up of negative linkage disequilibrium that cannot be
adequately captured by this model. We find accounting for genic selective
interference between segments fits human population genomic data slightly
better, suggesting levels of negative selection in humans are sufficient to
perturb the selection at other sites.

\section*{Results}
\subsection*{Theory}

Similar to previous work, our method characterizes the negative selection
process through its indirect effects on levels of neighboring pairwise
diversity at putatively neutral sites. Essential to this approach are accurate
theoretic models for how a selection process alters pairwise coalescence times
at sites some recombination distance away. The predominant model for how
negative (or purifying) selection reduces pairwise coalescence times at linked
sites is the classic background selection model
\parencite{Charlesworth1993-gb}. This simple and elegant model assumes new
deleterious mutations enter a region of $L$-basepairs at a rate of $\mu$ per
basepair per generation with a selection coefficient $s$ in heterozygotes that
is sufficiently deleterious that it is destined to loss. Under this setup, the
number of segregating deleterious mutations in the region reaches a stationary
Poisson distribution with rate $\nicefrac{U}{s}$ ($U=\mu L$), which is
precisely the equilibrium number under a deterministic mutation-selection
balance models \parencite{Kimura1966-bk,Crow1970-wj}. Modifications of this
basic model that allow for recombination between the segment under negative
selection and a focal neutral site \parencite{Nordborg1996-nq,Hudson1994-oh}
have been the basis of statistical methods that fit genome-wide patterns of
diversity to theoretic expectations \parencite{McVicker2009-ax,Hudson1995-xc}.
Furthermore, jointly incorporating both hitchhiking and background selection
have allowed for model comparison approaches that can estimate the relative
contributions of these two modes of selection
\parencite{Elyashiv2016-vt,Murphy2022-sj}.

These linked selection models predict how selection increases the rate
of coalescence at a neighboring neutral site some recombination distance away.

In general, these methods characterize selection processes by fitting the
relationship between observed pairwise diversity and expected levels according
to a theoretic model of linked selection. These linked selection models predict
how selection impacts the coalescence rate at some nearby site, which is
mediated by the strength of selection (e.g. the rate of new selected

For negative selection, the background
selection model predicts the expected pairwise coalescence time at some focal
neutral site near a region under mutation-selection balance. The increase in
coalescence rate

The impact of selection on linked neutral sites is mediated by the rate at
which selected mutations enter the population, the strength of selection, and
the recombination distance between

Both
hitchhiking and background selection models approximate the increased rate of
coalescence near sites under positive and negative selection as a reduction in
effective population size. Hitchhiking models

typically imagine a sweeping
beneficial allele creating

feature of linked selection theory is that the increased rate of coalescence at
sites neighboring regions under either positive or negative selection

Both positive
and negative selection increase the rate of coalescence at neighboring sites,
which both hitchhiking (XXX) and background selection (XXX) models

\begin{figure}[!htb]
  \centering
  \includegraphics{figures/figure1.pdf}
  \caption{}
  \label{fig:figure-1}
\end{figure}

@@

It is worth considering why this model performs well for models of negative
selection. The fitness variance equation, $V(x) = (U-2R)s$ tells us that under
models, the long-run variance is set by the balance of mutation and fixation.
Whether a particular basepair's behavior follows this equation depends upon
whether the long-run average is close to ``typical" for that basepair at a
given time. Good things fix so quickly, that we're left with a constant
churning of new variation
